Supplementary material: When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-03-01 doc/paper/body.tex
finished body, onto conclusions

  By accounting for photosynthesis changes in ecosystem conductance,
  with Eq. (\ref{et}) we have used recent results
  \citep[][]{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017} to derive
  ET explicitly as function of environmental variables and two
  plant-specific constants, the slope parameter ($g_1$), and uWUE. For
  the first time we have removed the implicit dependence of ET on
  itself through the stomatal conductance term, and we have also
  replaced the added complexity of a stomatal conductance reduction
  factor and a photosythesis model with a signle paramter (uWUE). Both
  the plant paramters reflect water conservation strategy.The
  slope parameter is related to the willingness of stomata to trade
  water for CO$_2$ and to keep stomata open (carbon cost in terhms of
  water)). uWUE is a semi-empirical ecosystem-scale constant related
  to how WUE changes with VPD (specifically $VPD^{-1/2}$). It is also
  roughly proportional to physical constants:

@@

With Equation \ref{et} we can take the partial derivative of ET with respect
to VPD to understand whether VPD drives or reduces ecosystem-scale ET:

@@

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefore, our
  derivation will be most applicable in times like the growing season
  (when we also expect uWUE to be most valid), when relatively high
  insolation induces instability and convective boundary layers, and
  we would expect the surface to be generally well coupled. An
  additional assumption in the formulation of uWUE \citep{Zhou_2014,
    Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
  that direct soil evaporation contributions to ET remain small
  relative to transpiration. Again, this should be more true during
  the growing season. The ratio of evaporation to transpiration may
  increase immediately after rainfall events due to high soil moisture
  and ponding, but VPD is generally low anyways during these
  times. However, some plant types allow for systematically larger
  contributions of evaporation in ET, particularly those with sparse
  canopies and smaller relative amounts of transpiration. We therefore
  might expect that the theory will be most applicable to forest PFTs,
  which will be most strongly coupled to the boundary layer due to
  larger surface roughness, and will also generally have the highest
  ratios of transpiration to evaporation. Finally, we assume that
  $g_1$ and uWUE are constant with respect to the conceptual VPD
  perturbation. Both quantities have been shown to be relatively
  constant with respect to changes in VPD \citep{Franks_2017,
    Zhou_2014}. These parameters will however vary with plant species
  and characteristics \citep[e.g. wood density, ][]{Lin_2015}, as well
  as environmental conditions including soil water content and
  temperature \citep{Lin_2015, Manzoni2013}. Exploring possible soil
  moisture (in)dependnce of the plant parameters (g$_1$ and uWUE) is
  particularly interesting because soil moisture only enters the
  partial derivative directly through these plant prameters. If the
  plant parameters are weak functions of soil moisture then the theory
  can be directly applied to a broader range of conceptual VPD
  scenarios, including observed compound events between high VPD and
  low soil moisture \citep{Zhou_2019}. Supplementary material for this
  manuscript explores the soil moisture dependence of the plant
  parameters, but excessive noise and inconsistencies preclude
  conclusions about the nature of any dependence. We provide it in
  case it is of interest to the reader, and to motivate future
  research.

@@

  To account for both the variability in plant parameters and the
  environment, we will systematically analyze how the ET response to
  VPD (Eq. (\ref{d_et})) varies. Equation (\ref{d_et}) includes a
  ``sign'' term that determines the sign of the response in addition
  to magnitude:

@@

Examining the sign term independent of the scaling term illuminates
the role of the two plant parameters, g$_1$ and uWUE, in determining
the degree to which the response is water conservative
($\frac{\partial \; ET}{\partial \; VPD} < 0$) or water intensive
($\frac{\partial \; ET}{\partial \; VPD} > 0$) (Figure \ref{sign})
. Higher g$_1$ and uWUE shift the curve towards increasing ET
responses with VPD (water intensive), and smaller g$_1$ values lead to
a higher non-linear VPD dependence of the response. g$_1$ and uWUE
should generally co-vary with $\lambda$. As $\lambda$ increases, g$_1$
should increase and uWUE should decrease. ** ellaborate **

@@

Examining the scaling term independent of the sign term shows how
aerodynamic conductance (communication between the atmosphere and the
surface) and temperature (controls the efficiency of energy conversion
to latent heat through $\Delta$), amplify or suppress the plant
response represented in the sign term (Figure \ref{scaing}). Both
lower temperatures and higher aerodynamic conductance lead to
amplified ET response to VPD, with the variability of aerodynamic
conductance resulting in a slightly higher variability of ET response
relative to temperature.

@@

Interpreting both the derivation and results of this manuscript hinges
on an understanding and appreciation for the usefulness of partial
derivatives for understanding the behavior of complex systems in a
simplifed framework. Our results are a conceptual tool for answering
the question, ``What is the ET response to VPD with all other
quantities held fixed?'' This is useful, and we would argue is
critical, for understanding complex responses when many quantities are
varying simultaneously \citep{Zhou_2019}. If we cannot understand the
contribution of one term to observed variability, independent of other
confounding factors, there is little hope of disentangling and
understanding the relative role of many covarying quantities in
determining observed variability.  Equation \ref{d_et} explicitly
provides an estimate of the ET response to VPD, given assumptions
about ecosystem plant characteristics through the parameters g$_1$ and
uWUE. Eq. (\ref{d_et}) is generic to external environmental factors
and the timescale over which a VPD response acts, and has a
straightforward conceptual interpretation. It provides the ET
response, given other quantities and parameters held fixed, subject to
the assumptions outlined in Sect. \ref{methods}. So, while the plant
parameters g$_1$ and uWUE may vary with plant species and
environmental conditions like soil moisture, we can still assess the
ET response to VPD for a given plant species, or for a given soil
moisture condition. ** talk about uWUE-SM, with supplementary here **?

However, given that g$_1$ and uWUE can vary, and they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}), it is
useful to examine how these quantities vary in previous literature,
framed by our results on $\frac{\partial \; ET}{\partial \; VPD}$. In
\citet{Zhou_2015} most plant types have uWUE similar to our median
value, under which the sign of the ET response depends on
VPD. However, crops, which we expect to prioritize carbon uptake over
water conservation, have a higher uWUE value closer to our maximum
value. This higher uWUE results in a higher likelihood for an
increasing ET response to VPD, which matches intuition given that we
expect crops to keep stomata open for access to carbon, at the cost of
increased water loss during high VPD. Shrubs, and to a lesser extent
grass, have an uWUE closer to the minimum uWUE. For these plant types,
we would then expect a decrease of ET in response to VPD. It is
important to note that, while \citet{Zhou_2015} did not examine the
role of soil moisture for within plant-type variability of uWUE, we
might expect some variation, especially in extreme cases when soil
water becomes a limiting factor. ** talk about uWUE-SM, with supplementary here **?

\citet{Medlyn_2017}'s results did not display the robust relationship
between plant type and and plant parameter (g$_1$) that was
demontrated by \citet{Zhou_2015} for uWUE. Given that there is a
robust relationship between g$_1$ and plant type in \citet{Lin_2015},
and that \citet{Zhou_2015}'s ecosystem-scale results rely on the same
underlying theory as \citet{Medlyn_2017}'s results, the ambiguous
results in \citet{Medlyn_2017} could be due to noise in ecosystem
scale observations, the consequences of imposing a model structure in
some of their g$_1$ esimations, or real ecosystem-scale within-plant
type variability (e.g. multiple species impacting the relationship
with the eddy-covariance footprint). These differences between
leaf-scale behavior and ecosystem scale behavior highlight the
importance of understanding how and why slope parameters show
increased variability at the ecosystem scale, as argued by
\citet{Medlyn_2017}. However, if we assume that there is some real
analogy between leaf scale results in \citet{Lin_2015} and ecosystem
scale behavior, and that some of the ambiguity in \citet{Medlyn_2017}
is due to model and/or observational error, then we can g$_1$-induced
ET response variability and plant characteristics, which are subject
to the strong caveat that we still do not fully understand g$_1$
behavior at the ecosystem scale. Extrapolating \citet{Lin_2015}'s
results (e.g. Fig. 2) to the ecosystem scale would reveal the
following relationships between g$_1$ and plant types:

@@

Additionally, the sensitivity of ET response to plant parameters
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf scale is not
analagous to the ecosystem scale must be understood. While the
derivation here was explicit about the assumption of a constant g$_1$
term with respect to the VPD perturbation, many sophisticated land
surface models and earth system models employing stomatal conductance
models make similar, and even stronger assumptions about the
stationarity of VPD slope terms \citep{Niu_2011, Franks_2017,
  Rogers_2017, Lawrence_2019}. Many models will assume a constant
g$_1$ term within a given plant functional type, and research derived
from these models often does not acknowledge limitations of this
assumption, or the difficulty in the literature with quantifying an
ecosystem scale slope parameter \citep{Medlyn_2017}. Given that
something so fundamental as the ET response to VPD varies strongly
with g$_1$, we must invest more in understanding g$_1$'s behavior at
the ecosystem scale, and the efficacy of models using g$_1$'s for
ecosystem-scale fluxes. Fortunately, our framework, and the novel
approach of using robust ecosystem-scale ratios (uWUE) to remove the
implicit dependence of stomatal conductance on ET, is flexible enough
to be applied to any stomatal conductance model that contains a
dependence on GPP.

So far we have discussed the role of different parameters in
modulating the ET response. However, the form of the ET response is
imposed by our choice of stomatal conductance model. We now examine
how the stomatal model choice, and specifically the exponent of the
VPD dependence in the stomatal conductance model, can impact the
general form of the ET-VPD relationship. There is a theoretical basis
for the square root VPD dependence in both the stomatal conductance
model and uWUE based on the assumption that stomata behave to
maximize carbon gain while minimizing water loss, which observations
also generally support \citep{Lloyd_1991, MEDLYN_2011, Lin_2015,
  Zhou_2014, Zhou_2015, Medlyn_2017}. However, some purely empirical
results that fit the exponent of the VPD dependence to data have shown
that it may vary slightly from 1/2, suggesting that stomata, as well
as ecosystem-scale quantities based on stomata theory, may not always
function optimally \citep{Zhou_2015, Lin_2018}. Specifically with
regards to uWUE, one would not expect that this ecosystem scale WUE
quantity will respond to VPD exactly analogously to stomata. Direct
soil evaporation's contributions to ET should shift the exponent of
the VPD dependence, especially at conditions of low GPP when we would
expect a systematically larger portion of direct soil evaporation
contributions to ET, because we would also expect lower amounts of
transpiration at low GPP. \citet{Zhou_2015}'s results corroborate
this: they found a mean empirically fit exponential VPD dependence of
0.55, varying slightly from the theoretically optimal value of of 0.5
for AmeriFlux sites. \citet{Lin_2018}'s results also show variance in
the empirical exponent of the VPD dependence of the stomatal
conductance model, but interpretation of this variance is more
difficult as \citet{Lin_2018} do not handle GPP/A dependence of
stomatal conductance in a directly analogous manner to the optimal
theory in \citet{MEDLYN_2011} and \citet{Medlyn_2017}. Regardless,
given that these recent results on the relationship between VPD, GPP,
and ET \citep{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017} form the
backbone of our analysis and are what allowed us to derive an explicit
ET expression for the first time (Eq. (\ref{et})), we will analyze if
and how assumptions about the exponent of the VPD dependence impacts
the shape of the ET-VPD dependence. This analysis is also important to
understand whether the choice of stomatal conductance model alters the
fundamental behavior of the ET-VPD relationships, as many commonly
used models utilize a VPD exponent other than the 1/2 suggested by
optimal theory \citep[e.g. ][ which uses an exponent of
1]{Leuning_1990}. It is useful to understand how the choice of model
imposes the functional form of the ET-VPD relationship, given its
fundamental importance for land-atmosphere interactions.

@@

The results presented in Sect. \ref{results} indicates that for a
given uWUE and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minimum in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms
(e.g. aerodynamic conductance, $\Delta$, $g_1$ and uWUE) are held
fixed. This result warrants further investigation, because to our
knowledge no earlier work has derived the theoretical ecosystem-scale
relationship between ET and VPD while controlling for other
environmental conditions. In particular there is an apparent lack of
consensus over whether the shape of the ET-VPD curve should be concave
upward (our result) or concave downward in the absence of dramatic
water stress. Given that understanding the ET-VPD relationship of the
one-way plant response is fundamental to hypothesizing about any
feedbacks between the land surface and the atmosphere, we analyze why
our derived ET-VPD relationship is concave upward, particularly with
respect to the exponent of VPD dependence in uWUE and the Medlyn
unified stomatal conductance model, as other models and empiracle
results have suggested exponents varying from 1/2 \citep{Leuning_1990,
  Zhou_2015, Lin_2018}.

By introducing $n$, the exponent of VPD in uWUE, and $m$, the exponent
of VPD in the stomatal conductance model, we can free our theory from
assumptions about VPD dependence:
% \begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g*}{VPD^m}\right) \frac{*WUE \; ET}{c_a \; VPD^n},
    \label{m_n}
  \end{equation}
% \end{linenomath*}
where:
\[*WUE = \frac{GPP}{ET}VPD^n,\] and $g*$ is a generic slope parameter
of units $VPD^m$. To determine how the exponent $n$ and $m$ alter the
shape of the ET-VPD dependence we find the roots of the second
derivative of ET, using Eq. (\ref{m_n}) for stomatal conductance
($g_s$), with respect to VPD: \input{d2_solutions.tex} With this
result we have defined the family of curves separating concave up from
concave down ET solutions (Fig. \ref{concave}). These curves are
only functions of the exponent of the VPD dependence and a quantity we
call non-dimensional VPD ($VPD^m/g*$). Several important
relations reveal themselves from Eq. (\ref{curves}):

@@

While it is possible that in the future some other form of
VPD dependence is derived, at present \cite{MEDLYN_2011} and
\cite{Zhou_2014} firmly established n=m=1/2 as the most likely
candidate given current theory and empirical data. Additionally, we
argue that a concave up result matches physical intuition more than a
concave down result. Plants must maintain nutrient and sugar transport
through the phloem and xylem. To accomplish this, stomata must remain
slightly open \citep{De_2013, Nikinmaa_2013, Ryan_2014}. Furthermore,
even if complete stomatal closure were possible, cuticular water loss
and [at the ecosystem-scale] direct soil evaporation are still sources
of ET which increase with VPD, independent of stomatal
closure ** cite katul??? **. Therefore, in the limit as VPD becomes large and we assume
plants are exercising all strategies to reduce ET, any further increase
in VPD should result in an increase in ET through cuticular water loss
and/or direct soil evaporation. This inevitable transition from
conditions when stomata respond strongly to VPD to conditions when
stomata response is asymptoting towards full closure would cause a
concave up ET-VPD curve, which is matched by our theory. In short,
plant response becomes more limited as VPD increases, while
atmospheric demand monotonically increases with VPD, leading to the
 result that atmospheric demand dominates plant response when
atmospheric demand is high.

This analysis allows us to understand the theoretical shape of the ET
response to VPD with environmental conditions held
fixed. Accomplishing this with purely statistical methods applied to
flux observations would be very difficult, given the relatively fast
time scale of plant response and the non-stationarity of [solar
forced] environmental conditions over the relatively coarse (half
hourly) flux estimates (which is required to obtain robust
eddy-covariance statistics). Our results on the shape of the ET-VPD
curve with environmental conditions held fixed can be built upon with
future work examining how changes in VPD and environmental conditions
(e.g. soil water storage) feedback upon one another. In the soil water
storage example, over very long time scales extremely high VPD
perturbations coupled with no precipitation could result in decreases
in soil water storage such that water becomes limiting. This could be
represented by an extension of our framework in which uWUE is allowed
to change with changing SWC. Here, we focus our results by framing
theme with previous literature to build baseline intuition for ET-VPD
dependence.

>>>>

  By accounting for photosynthesis changes in ecosystem conductance,
  with Eq. (\ref{et}) we have used recent results
  \citep[][]{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017} to derive
  ET explicitly as function of environmental variables and two
  plant-specific constants, the slope parameter ($g_1$), and uWUE. For
  the first time we have removed the implicit dependence of ET on
  itself through the stomatal conductance term, and we have also
  replaced the added complexity of a stomatal conductance reduction
  factor and a photosynthesis model with a single parameter (uWUE). Both
  the plant parameters reflect water conservation strategy.The
  slope parameter is related to the willingness of stomata to trade
  water for CO$_2$ and to keep stomata open (carbon cost in terms of
  water)). uWUE is a semi-empirical ecosystem-scale constant related
  to how WUE changes with VPD (specifically $VPD^{-1/2}$). It is also
  roughly proportional to physical constants:

@@

With Eq. (\ref{et}) we can take the partial derivative of ET with respect
to VPD to understand whether VPD drives or reduces ecosystem-scale ET:

@@

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefore, our
  derivation will be most applicable in times like the growing season
  (when we also expect uWUE to be most valid), when relatively high
  insolation induces instability and convective boundary layers, and
  we would expect the surface to be generally well coupled. An
  additional assumption in the formulation of uWUE \citep{Zhou_2014,
    Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
  that direct soil evaporation contributions to ET remain small
  relative to transpiration. Again, this should be more true during
  the growing season. The ratio of evaporation to transpiration may
  increase immediately after rainfall events due to high soil moisture
  and ponding, but VPD is generally low anyways during these
  times. However, some plant types allow for systematically larger
  contributions of evaporation in ET, particularly those with sparse
  canopies and smaller relative amounts of transpiration. We therefore
  might expect that the theory will be most applicable to forest PFTs,
  which will be most strongly coupled to the boundary layer due to
  larger surface roughness, and will also generally have the highest
  ratios of transpiration to evaporation. Finally, we assume that
  $g_1$ and uWUE are constant with respect to the conceptual VPD
  perturbation. Both quantities have been shown to be relatively
  constant with respect to changes in VPD \citep{Franks_2017,
    Zhou_2014}. These parameters will however vary with plant species
  and characteristics \citep[e.g. wood density, ][]{Lin_2015}, as well
  as environmental conditions including soil water content and
  temperature \citep{Lin_2015, Manzoni2013}. Exploring possible soil
  moisture (in)dependence of the plant parameters (g$_1$ and uWUE) is
  particularly interesting because soil moisture only enters the
  partial derivative directly through these plant parameters. If the
  plant parameters are weak functions of soil moisture then the theory
  can be directly applied to a broader range of conceptual VPD
  scenarios, including observed compound events between high VPD and
  low soil moisture \citep{Zhou_2019}. Supplementary material for this
  manuscript explores the soil moisture dependence of the plant
  parameters, but excessive noise and inconsistencies preclude
  conclusions about the nature of any dependence. We provide it in
  case it is of interest to the reader, and to motivate future
  research.

@@

  To account for both the variability in plant parameters and the
  environment, we will systematically analyze how the ET response to
  VPD (Eq. (\ref{d_et})) varies. Eq. (\ref{d_et}) includes a
  ``sign'' term that determines the sign of the response in addition
  to magnitude:

@@

Examining the sign term independent of the scaling term illuminates
the role of the two plant parameters, g$_1$ and uWUE, in determining
the degree to which the response is water conservative
($\frac{\partial \; ET}{\partial \; VPD} < 0$) or water intensive
($\frac{\partial \; ET}{\partial \; VPD} > 0$) (Figure \ref{sign})
. Higher g$_1$ and uWUE shift the curve towards increasing ET
responses with VPD (water intensive), and smaller g$_1$ values lead to
a higher non-linear VPD dependence of the response. However care
should be exercised when interpreting the range of ET responses,
because g$_1$ and uWUE should generally be anti-correlated due to
their dependencies on $\lambda$. As $\lambda$ increases, g$_1$ should
increase and uWUE should decrease. Because high and low values of uWUE
and g$_1$ were determined independently from each other in
previous literature, plant types with anomalously high or low values
for \textit{both} g$_1$ and uWUE should be relatively rare.

@@

Examining the scaling term independent of the sign term shows how
aerodynamic conductance (communication between the atmosphere and the
surface) and temperature (controls the efficiency of energy conversion
to latent heat through $\Delta$), amplify or suppress the plant
response represented in the sign term (Figure \ref{scaling}). Both
lower temperatures and higher aerodynamic conductance lead to
amplified ET response to VPD, with the variability of aerodynamic
conductance resulting in a slightly higher variability of ET response
relative to temperature.

@@

Interpreting both the derivation and results of this manuscript hinges
on an understanding and appreciation for the usefulness of partial
derivatives for understanding the behavior of complex systems in a
simplified framework. Our results are a conceptual tool for answering
the question, ``What is the ET response to VPD with all other
quantities held fixed?'' This is useful, and we would argue is
critical, for understanding complex responses when many quantities are
varying simultaneously \citep{Zhou_2019}. If we cannot understand the
contribution of one term to observed variability, independent of other
confounding factors, there is little hope of disentangling and
understanding the relative role of many co-varying quantities in
determining observed variability.  Eq. (\ref{d_et}) explicitly
provides an estimate of the ET response to VPD, given assumptions
about ecosystem plant characteristics through the parameters g$_1$ and
uWUE. Eq. (\ref{d_et}) is generic to external environmental factors
and the timescale over which a VPD response acts, and has a
straightforward conceptual interpretation. It provides the ET
response, given other quantities and parameters held fixed, subject to
the assumptions outlined in Sect. \ref{methods}. So, while the plant
parameters g$_1$ and uWUE may vary with plant species and
environmental conditions like soil moisture, we can still assess the
ET response to VPD for a given ecosystem state, or for a given soil
moisture condition.

However, given that g$_1$ and uWUE can vary, and they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}), it is
useful to examine how these quantities vary in previous literature,
framed by our results on $\frac{\partial \; ET}{\partial \; VPD}$. In
\citet{Zhou_2015} most plant types have uWUE similar to our median
value, under which the sign of the ET response depends on
VPD. However, crops, which we expect to prioritize carbon uptake over
water conservation, have a higher uWUE value closer to our maximum
value. This higher uWUE results in a higher likelihood for an
increasing ET response to VPD, which matches intuition given that we
expect crops to keep stomata open for access to carbon, at the cost of
increased water loss during high VPD. Shrubs, and to a lesser extent
grass, have an uWUE closer to the minimum uWUE. For these plant types,
we would then expect a decrease of ET in response to VPD. It is
important to note that, while \citet{Zhou_2015} did not examine the
role of soil moisture for within plant-type variability of uWUE, we
might expect some variation, especially in extreme cases when soil
water becomes a limiting factor (see supplementary material).

\citet{Medlyn_2017}'s results did not display the robust relationship
between plant type and and plant parameter (g$_1$) that was
demonstrated by \citet{Zhou_2015} for uWUE. Given that there is a
robust relationship between g$_1$ and plant type in \citet{Lin_2015},
and that \citet{Zhou_2015}'s ecosystem-scale results rely on the same
underlying theory as \citet{Medlyn_2017}'s results, the ambiguous
results in \citet{Medlyn_2017} could be due to noise in ecosystem
scale observations, the consequences of imposing a model structure in
some of their g$_1$ estimations, or real ecosystem-scale within-plant
type variability (e.g. multiple species impacting the relationship
within the eddy-covariance footprint). These differences between
leaf-scale behavior and ecosystem scale behavior highlight the
importance of understanding how and why slope parameters show
increased variability at the ecosystem scale, as argued by
\citet{Medlyn_2017}. However, if we assume that there is some real
analogy between leaf scale results in \citet{Lin_2015} and ecosystem
scale behavior, and that some of the ambiguity in \citet{Medlyn_2017}
is due to model and/or observational error, then we can use the
relationships between g$_1$ and plant characteristics defined by \citet{Lin_2015} to
frame the ET response to VPD. This is subject
to the strong caveat that we still do not fully understand g$_1$
behavior at the ecosystem scale. Extrapolating \citet{Lin_2015}'s
results (e.g. Fig. 2) to the ecosystem scale would reveal the
following relationships between g$_1$ and plant types:

@@

Additionally, the sensitivity of ET response to plant parameters
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf scale is not
analogous to the ecosystem scale must be understood. While the
derivation here was explicit about the assumption of a constant g$_1$
term with respect to the VPD perturbation, many sophisticated land
surface models and earth system models employing stomatal conductance
models make similar assumptions about the stationarity of VPD slope
terms \citep{Niu_2011, Franks_2017, Rogers_2017, Lawrence_2019}. Many
models will assume a constant VPD slope term (e.g. g$_1$) term within
a given plant functional type, and research derived from these models
often does not acknowledge limitations of this assumption, or the
difficulty in the literature with quantifying an ecosystem scale slope
parameter \citep{Medlyn_2017}. Given that something so fundamental as
the ET response to VPD varies strongly with g$_1$, we must invest more
in understanding g$_1$'s behavior at the ecosystem scale, and the
efficacy of models using constant slope terms for ecosystem-scale
fluxes. Fortunately, our framework, and the new approach of using
robust ecosystem-scale ratios (uWUE) to remove the implicit dependence
of stomatal conductance on ET, is flexible enough to be applied to any
stomatal conductance model that contains a dependence on GPP.

So far we have discussed the role of different parameters in
modulating the ET response. However, the form of the ET response is
imposed by our choice of stomatal conductance model. We now examine
how the stomatal model choice, and specifically the exponent of the
VPD dependence in the stomatal conductance model, can impact the
general form of the ET-VPD relationship. There is a theoretical basis
for the square root VPD dependence in both the stomatal conductance
model and uWUE based on the assumption that stomata behave to maximize
carbon gain while minimizing water loss, which observations also
generally support \citep{Lloyd_1991, MEDLYN_2011, Lin_2015, Zhou_2014,
  Zhou_2015, Medlyn_2017}. However, some purely empirical results that
fit the exponent of the VPD dependence to data have shown that it may
vary slightly from 1/2, suggesting that stomata, as well as
ecosystem-scale quantities based on stomata theory, may not always
function optimally \citep{Zhou_2015, Lin_2018}. Specifically with
regards to uWUE, one would not expect that this ecosystem scale WUE
quantity will respond to VPD exactly analogously to stomata. Direct
soil evaporation's contributions to ET should shift the exponent of
the VPD dependence. \citet{Zhou_2015}'s results corroborate this: they
found a mean empirically fit exponential VPD dependence of 0.55,
varying slightly from the theoretically optimal value of of 0.5 for
AmeriFlux sites. \citet{Lin_2018}'s results also show variance in the
empirical exponent of the VPD dependence in the stomatal conductance
model, but interpretation of this variance is more difficult as
\citet{Lin_2018} do not handle the GPP dependence of stomatal
conductance in a directly analogous manner to the optimal theory in
\citet{MEDLYN_2011} and \citet{Medlyn_2017}. Regardless, given that
these recent results on the relationship between VPD, GPP, and ET
\citep{MEDLYN_2011, Zhou_2014, Zhou_2015, Medlyn_2017} form the
backbone of our analysis and are what allowed us to derive an explicit
ET expression for the first time (Eq. (\ref{et})), we will analyze if
and how assumptions about the exponent of the VPD dependence impact
the shape of the ET-VPD dependence. This analysis is also important to
understand whether the choice of stomatal conductance model alters the
fundamental behavior of the ET-VPD relationships, as many commonly
used models utilize a VPD exponent other than the 1/2 suggested by
optimal theory \citep[e.g. ][ which uses an exponent of
1]{Leuning_1990}. It is useful to understand how the choice of model
imposes the functional form of the ET-VPD relationship, given its
fundamental importance for land-atmosphere interactions.

@@

The results presented in Sect. \ref{results} indicates that for a
given uWUE and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minimum in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms
(e.g. aerodynamic conductance, $\Delta$, $g_1$ and uWUE) are held
fixed. This result warrants further investigation, because to our
knowledge no earlier work has derived the theoretical ecosystem-scale
relationship between ET and VPD in an energy balance framework, while
controlling for other environmental conditions. In particular there is
an apparent lack of consensus over whether the shape of the ET-VPD
curve should be concave upward (our result) or concave downward in the
absence of dramatic water stress. Given that understanding the ET-VPD
relationship of the one-way plant response is fundamental to
hypothesizing about any feedbacks between the land surface and the
atmosphere, we analyze why our derived ET-VPD relationship is concave
upward, particularly with respect to the exponent of VPD dependence in
uWUE and the Medlyn unified stomatal conductance model, as other
models and empirical results have suggested exponents varying from 1/2
\citep{Leuning_1990, Zhou_2015, Lin_2018}.

By introducing $n$, the exponent of VPD in uWUE, and $m$, the exponent
of VPD in the stomatal conductance model, we can free our theory from
assumptions about VPD dependence:
% \begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g*}{VPD^m}\right) \frac{*WUE \; ET}{c_a \; VPD^n},
    \label{m_n}
  \end{equation}
% \end{linenomath*}
where:
\[*WUE = \frac{GPP}{ET}VPD^n,\] and $g*$ is a generic slope parameter
of units $VPD^m$. To determine how the exponent $n$ and $m$ alter the
shape of the ET-VPD dependence we find the roots of the second
derivative of ET, using Eq. (\ref{m_n}) for stomatal conductance
($g_s$), with respect to VPD:

\input{d2_solutions.tex}

With this
result we have defined the family of curves separating concave up from
concave down ET solutions (Fig. \ref{concave}). These curves are
only functions of the exponent of the VPD dependence and a quantity we
call non-dimensional VPD ($VPD^m/g*$). Several important
relations reveal themselves from Eq. (\ref{curves}):

@@

While it is possible that in the future some other form of VPD
dependence is derived, at present \cite{MEDLYN_2011} and
\cite{Zhou_2014} established n=m=1/2 as the most likely candidate
given current theory and empirical data. Additionally, we argue that a
concave up result matches physical intuition more than a concave down
result. Plants must maintain nutrient and sugar transport through the
phloem and xylem. To accomplish this, stomata must remain slightly
open \citep{De_2013, Nikinmaa_2013, Ryan_2014}. Furthermore, even if
complete stomatal closure were possible, cuticular water loss and [at
the ecosystem-scale] direct soil evaporation are still sources of ET
which increase with VPD, independent of stomatal closure. Therefore,
in the limit as VPD becomes large and we assume plants are exercising
all strategies to reduce ET, any further increase in VPD should result
in an increase in ET through cuticular water loss and/or direct soil
evaporation. This inevitable transition from conditions when stomata
respond strongly to VPD to conditions when stomata response is
asymptoting towards full closure would cause a concave up ET-VPD
curve, which is matched by our theory. In short, plant response
becomes more limited as VPD increases, while atmospheric demand
monotonically increases with VPD, leading to the result that
atmospheric demand dominates plant response when atmospheric demand is
high.

This analysis allows us to understand the theoretical shape of the ET
response to VPD with environmental conditions held
fixed. Accomplishing this with purely statistical methods applied to
flux observations would be very difficult, given the relatively fast
time scale of plant response and the non-stationarity of [solar
forced] environmental conditions over the relatively coarse (half
hourly) flux estimates (which is required to obtain robust
eddy-covariance statistics). Our results on the shape of the ET-VPD
curve with environmental conditions held fixed can be built upon with
future work examining how changes in VPD and environmental conditions
(e.g. soil water storage) feedback upon one another. In the soil water
storage example, over very long time scales extremely high VPD
perturbations coupled with no precipitation could result in decreases
in soil water storage such that water becomes limiting. This could be
represented by an extension of our framework in which $\lambda$ is
allowed to decrease with decreasing soil water, increasing uWUE and
decreasing g$_1$. Here, we focus our results by framing them with
previous literature to build baseline intuition for ET-VPD dependence.
